---
title: "Class11"
author: "Brianna Smith"
format: pdf
---

## Section 1 Proportion of G/G in a population

Downloaded a CSV file from Ensemble 

Here we read this CSV file

```{r}
mxl <- read.csv("mxldata.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)
```





##Section 4: Population Scale Analysis[HOMEWORK]

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
summary(expr$exp)
```

```{r}
nrow(expr)
```

sample size for each genotype:
```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Let's make a boxplot:

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
```

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

From the blox plot, we can see that A|A genotype seems to be more associated with higher expression than the G|G genotype. The G|G genotype is associated with having reduced expression of the gene.









